### 2019-08-07 paper/paper.tex
revised the methods a little

Alternatively, reaction-diffusion dynamics could generate a Voronoi-like tesselation in the cortex without the need for pre-arranged centres, by amplifying characteristic modes in the noisy initial distribution of bundled axon branches, as a net effect of short-range cooperative and longer-range competitive interactions between cells. According to this model the barrel pattern would be determined by the relative strength of these interactions and by the shape of the cortical field boundary. However, reaction-diffusion dynamics alone are unlikely to be able to account for the precise co-registration and topographic connectivity between emergent domains in thalamus and cortex, the non-uniform sizes and spacing of the barrels to form rows versus arcs, and the contribution of gene expression gradients.

@@

\cite{Karbowski2004} developed what remains, to the best of our knowledge, the only dynamical model of cortical arealization capable of generating predictions about the influence of signalling molecules on the size, shape, and topological organization of emergent cortical fields. This model describes competition between thalamocortical projections for territory along a one-dimensional string of cortical cells, e.g., along the anterior-posterior axis. The equations of the model define how the density of connections, $c$, and the density of axon branches, $a$, interact at time $t$ at each cortical location $x$, for $N$ thalamocortical projections indexed by $i$. The tesselation of the barrels that we investigate is a two-dimensional structure, hence we first extended the original model for simulation in two spatial dimensions, and describe it here as such (i.e., suibstituting $x$ for $\mb{x}$ and $\frac{d}{dx}$ for $\nabla$).

The model was derived from the assumption that the rates of synaptogenesis and thalamocortical axon branching are reciprocally coupled, resulting in an expression for the change in $c_i$ as a function of $a_i$, and for the change in $a_i$ as a function of $c_i$. Synaptogenesis is modelled  as

\begin{equation} \label{eq:Karb2DExt_dc}
\frac{\partial c_i(\mb{x},t)}{\partial t} =-\alpha c_i(\mb{x},t) +\beta  \left(1 - \sum_{j=1}^{N} c_j(\mb{x}, t)\right).[a_i(\mb{x},t)]^k.
\end{equation}

Accordingly when the final term evaluates to zero, the connection density decays exponentially at a rate given by the decay constant $\alpha$. This occurs, at a given location, when the sum of connections is one. Otherwise, the connection density is assumed to vary non-linearly ($k>1$) with changes in the axon branching density.

The change in density of axon branches is assumed to depend also on the influence of guidance molecules and on the density of branches at adjacent cortical locations:

%
\begin{equation} \label{eq:Karb2DExt_da}
\frac{\partial a_i(\mb{x},t)}{\partial t} = \nabla\vcdot\left(D \nabla a_i(\mb{x},t)-a_i(\mb{x},t)\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i(\mb{x},t)}{\partial t} + \chi
\end{equation}

The first term on the right describes the spatial gradient (indicated by $\nabla$) of a quantity referred to in models of heat diffusion as the `flux' (enclosed by parentheses). The central idea in such models is that the change in temperature (over time) at a given point on a surface is proportional to the gradient in the temperature (over space) at that location. Substituting heat for the number of branches, axons from a given thalamic projection are thus assumed to branch laterally across the cortical sheet at a rate given by the diffusion constant $D$.

Lateral branching is augmented by a measure of the local alignment between the density of branches and the gradient in the concentration of each of $M$ distinct molecular signals, $\rho(\mb{x})$. The influence of a given signal (indexed by $j$) on a given projection (indexed by $i$), is determined by the parameter $\gamma_i,j$. Interactions for which this value is positive encourage axon branching in the direction in which the concentration of the signal is increasing, whereas negative interactions encourage branching in the opposite direction. Hence the axons of a given projection are assumed to branch towards either higher or lower concentrations of each signalling molecule.

The second term on the right of equation 2 represents an assumption that the number of axon branches decreases when the number of connections being made increases, and vice versa, i.e., that branches are required to form connections. Note that we introduce the term $\chi=0$ here, so that we might subsequently redefine $\chi$ in order to introduce the competition between thalamic projections that we found to be necessary for the emergence of a Voronoi tesselation.

Two points are important to emphasise here. First, notice that the only difference between thalamic projections is the $M$ interaction strengths ($\gamma$) that are chosen to define each projection. Hence any (reliable) differences between the size, shape, and location of simulated cortical fields are due to differences in the choice of these values only. Second, the formulation of the model is at this point entirely local, in the sense that simulating the dynamics does not require any information to be communicated from one cortical cell to any but its immediate neighbours.

In addition to the choice of initial conditions (for the fields $c(\mb{x},t=0)$ and $a(\mb{x},t=0)$) and the axon guidance fields $\rho(\mb{x})$, the free parameters of the model are the number of thalamocortical projections ($N$), constants $\alpha$, $\beta$, $k$, and $D$, and the $N\times M$ interaction strengths, $\gamma$.

Equations 1 and 2 were solved on a hexagonal lattice of grid elements represent the cortical sheet ($\mb{x}$), under the assumption that axon branches cannot extend beyond the boundary of the tissue, i.e., $\mb{J}_i(\mb{x},t) \bigg\rvert_{boundary} = 0$. We developed software tools for imposing arbitrary boundary shapes, specified in a scalable vector graphics (svg) file format, on the lattice. The dynamics were integrated using fourth-order Runge-Kutta for ordinary differential equation 1, and using a {\color{red} finite difference method} for the flux term (see Supplementary Materials).

>>>>

Alternatively, reaction-diffusion dynamics could generate a Voronoi-like tesselation in the cortex without the need for pre-arranged centres, by amplifying characteristic modes in the noisy initial distribution of bundled axon branches, as a net effect of short-range cooperative and longer-range competitive interactions between cells. According to this model the barrel pattern would be determined by the relative strength of these interactions and by the shape of the cortical field boundary. However, reaction-diffusion dynamics alone are unlikely to be able to account for the precise co-registration and topographic connectivity between emergent domains in thalamus and cortex, the non-uniform sizes and spacing of the barrels to form rows versus arcs, and the influence of gene expression gradients.

@@

\cite{Karbowski2004} developed what remains, to the best of our knowledge, the only dynamical model of cortical arealization capable of generating predictions about the influence of signalling molecules on the size, shape, and topological organization of emergent cortical fields. This model describes competition between thalamocortical projections for territory along a one-dimensional string of cortical cells, e.g., along an anterior-posterior axis. The equations of the model define how the density of connections, $c$, and the density of axon branches, $a$, interact at time $t$ at each cortical location $x$, for $N$ thalamocortical projections indexed by $i$. The Voronoi tesselation used to describe barrel patterning is a two-dimensional structure, hence we first extended the original model for simulation on a two-dimensional cortical sheet, and describe it here as such (i.e., substituting $x$ for the two-dimensional position vector $\mb{x}$ and substituting terms of the form $\frac{d}{dx}$ for the two-dimensional spatial operator $\nabla$).

The original model was derived from the assumption that the rates at which thalamocortical synapses form and thalamocortical axons branch are reciprocally coupled, resulting in an expression for the change in $c_i$ as a function of $a_i$, and for the change in $a_i$ as a function of $c_i$. Synaptogenesis is modelled  as

\begin{equation} \label{eq:Karb2DExt_dc}
\frac{\partial c_i(\mb{x},t)}{\partial t} =-\alpha c_i(\mb{x},t) +\beta  \left(1 - \sum_{j=1}^{N} c_j(\mb{x}, t)\right)[a_i(\mb{x},t)]^k.
\end{equation}

Accordingly, when the final term evaluates to zero, the density of thalamocortical connections decays exponentially at a rate given by the decay constant $\alpha$. This occurs where the density of connections over all thalamic projections sum to one. Otherwise, the connection density is assumed to vary non-linearly ($k>1$) with changes in the axon branching density.

Changes in axon brtanching density are assumed to depend also on the influence of guidance molecules and on the density of branches at adjacent cortical locations:

%
\begin{equation} \label{eq:Karb2DExt_da}
\frac{\partial a_i(\mb{x},t)}{\partial t} = \nabla\vcdot\left(D \nabla a_i(\mb{x},t)-a_i(\mb{x},t)\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i(\mb{x},t)}{\partial t} + \chi.
\end{equation}

The first term on the right describes the spatial gradient (indicated by $\nabla\vcdot$) of a quantity referred to in models of heat diffusion as the `flux' (the term enclosed by parentheses). The central idea in heat diffusion models is that the change in temperature (over time) at a given location is proportional to the gradient in the temperature (over space) at that location. Substituting heat for the branching density, axons from a given thalamic projection are thus assumed to branch laterally across the cortical sheet at a rate given by the diffusion constant $D$. Lateral branching is augmented by a measure of the local alignment between the branching density and the gradient in the concentration of each of $M$ distinct molecular signals, $\rho(\mb{x})$. The influence of a given signal (indexed by $j$) on a given thalamic projection (indexed by $i$), is determined by the parameter $\gamma_{i,j}$. Interactions for which this value is positive encourage axon branching in the direction in which the concentration of the signal is increasing, whereas negative interactions encourage branching in the opposite direction. Hence the axons of a given thalamic projection are assumed to branch towards either higher or lower concentrations of each signalling molecule.

The second term on the right of equation 2 represents an assumption that the axon branching density decreases when the number of connections being made increases, and vice versa, i.e., that branches are required to form synaptic connections. Note that we introduce the term $\chi=0$ here, so that we might subsequently redefine $\chi$ to introduce the competition between thalamic projections that we will show to be necessary for the emergence of a Voronoi tesselation (see Results).

Two points are important to emphasise here. First, notice that the only difference between thalamic projections is the interaction strengths, $\gamma$. Hence any differences between the size, shape, and location of cortical fields that reliably emerge from the dynamics of the model must be due to differences in the choice of these values only. Second, the formulation of the model is at this point entirely local, in the sense that simulating the dynamics requires no information to be communicated from cortical locations to any but the immediately adjacent locations.

In addition to the choice of initial conditions (for the fields $c(\mb{x},t=0)$ and $a(\mb{x},t=0)$) and the axon guidance fields $\rho(\mb{x})$, the free parameters of the model are the number of thalamocortical projections, $N$, constants $\alpha$, $\beta$, $k$, and $D$, and the $N\times M$ interaction strengths, $\gamma$.

The dynamics of the model (equations 1 and 2) were solved on a hexagonal lattice of grid elements representing the cortical sheet, under the assumption that axon branches cannot extend beyond the boundary of the sheet, i.e., $\mb{J}_i(\mb{x},t) \bigg\rvert_{boundary} = 0$. We developed software tools for imposing arbitrary boundary shapes, specified in a scalable vector graphics (.svg) file format, on the lattice. The dynamics were numerically integrated using fourth-order Runge-Kutta, and using a {\color{red} finite difference method} to compute the flux term (see Supplementary Materials).
